::: {.callout-important collapse="true"}
## Data structure
The data should be in a **long format** (also known as tidy data), where each row is an observation and each column is a variable (@fig-format). If your data is not already structured this way, reshape it manually in a spreadsheet program or in R using the `pivot_longer()` function from the `tidyr` package.


::: {layout="[ 0.4, -0.1, 0.4 ]" #fig-format}

:::: {#first-column}
```{r}
#| echo: false
possums <- readxl::read_excel("possum_bw.xlsx")
possums |>
  head(8) |>
  gt() |>
  opt_stylize(style = 5, color = "red")
```
:::

::: {#second-column}
```{r}
#| echo: false
possums |>
  group_by(Sex) |>
  mutate(row = row_number()) |>
  pivot_wider(names_from = Sex, values_from = BW) |>
  select(-row) |>
  head(4) |>
  gt() |>
  opt_stylize(style = 5, color = "red")
```
:::

Data should be in long format (left) where each row is an observation and each column is a variable. This is the preferred format for most statistical software. Wide format (right) is also common, but may require additional steps to analyse or visualise in some instances.
::::
:::